---
title: "CHIS Exercise"
author: "Michael Moran"
date: "July 12, 2017"
output: html_document
---

Part 1: Exploring Data

# Explore the dataset with summary and str

```{r, eval=FALSE, include=TRUE}
summary(adult)
str(adult)
```

# Age histogram

```{r, eval=FALSE, include=TRUE}
ggplot(adult, aes (x = SRAGE_P)) +
  geom_histogram()
```

# BMI histogram

```{r, eval=FALSE, include=TRUE}
ggplot(adult, aes (x = BMI_P)) +
  geom_histogram()
```


# Age colored by BMI, default binwidth

```{r, eval=FALSE, include=TRUE}
ggplot(adult, aes (x = SRAGE_P, fill= factor(RBMI))) +
  geom_histogram(binwidth = 1)
```

Part 2: Data Cleaning

# Remove individual aboves 84

```{r, eval=FALSE, include=TRUE}
adult <- adult[adult$SRAGE_P <= 84, ]
```

# Remove individuals with a BMI below 16 and above or equal to 52

```{r, eval=FALSE, include=TRUE}
adult <- adult[adult$BMI_P >= 16 & adult$BMI_P < 52, ]
```

# Relabel the race variable:

```{r, eval=FALSE, include=TRUE}
adult$RACEHPR2 <- factor(adult$RACEHPR2, labels = c("Latino", "Asian", "African American", "White"))
```

# Relabel the BMI categories variable:

```{r, eval=FALSE, include=TRUE}
adult$RBMI <- factor(adult$RBMI, labels = c("Under-weight", "Normal-weight", "Over-weight", "Obese"))
```

Part 3: Multiple Histograms

# The dataset adult is available

# The color scale used in the plot

```{r, eval=FALSE, include=TRUE}
BMI_fill <- scale_fill_brewer("BMI Category", palette = "Reds")
```

# Theme to fix category display in faceted plot

```{r, eval=FALSE, include=TRUE}
fix_strips <- theme(strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.1, size = 14),
                    strip.background = element_blank(),
                    legend.position = "none")
```

# Histogram, add BMI_fill and customizations

```{r, eval=FALSE, include=TRUE}
ggplot(adult, aes (x = SRAGE_P, fill= factor(RBMI))) +
  geom_histogram(binwidth = 1) +
  fix_strips + BMI_fill + facet_grid(RBMI ~.) + theme_classic()
```

Part 4: Alternatives

# Plot 1 - Count histogram

```{r, eval=FALSE, include=TRUE}
ggplot(adult, aes (x = SRAGE_P, fill= factor(RBMI))) +
  geom_histogram(binwidth = 1) +
  BMI_fill
```

# Plot 2 - Density histogram

```{r, eval=FALSE, include=TRUE}
ggplot(adult, aes (x = SRAGE_P, fill= factor(RBMI))) +
  geom_histogram(aes(y = ..density..), binwidth = 1) +
  BMI_fill
```

# Plot 3 - Faceted count histogram

```{r, eval=FALSE, include=TRUE}
ggplot(adult, aes (x = SRAGE_P, fill= factor(RBMI))) +
  geom_histogram(binwidth = 1) +
  BMI_fill +
  facet_grid(RBMI ~ .)
```

# Plot 4 - Faceted density histogram

```{r, eval=FALSE, include=TRUE}
ggplot(adult, aes(x = SRAGE_P, fill= factor(RBMI))) +
  geom_histogram(aes(y = ..density..), binwidth = 1) +
  BMI_fill +
  facet_grid(RBMI ~ .)
```

# Plot 5 - Density histogram with position = "fill"

```{r, eval=FALSE, include=TRUE}
ggplot(adult, aes(x = SRAGE_P, fill= factor(RBMI))) +
  geom_histogram(aes(y = ..density..), binwidth = 1, position = "fill") +
  BMI_fill
```

# Plot 6 - The accurate histogram

```{r, eval=FALSE, include=TRUE}
ggplot(adult, aes (x = SRAGE_P, fill= factor(RBMI))) +
  geom_histogram(aes(y = ..count../sum(..count..)), binwidth = 1, position = "fill") +
  BMI_fill
```

Part 5: Do Things Manually

# An attempt to facet the accurate frequency histogram from before (failed)

```{r, eval=FALSE, include=TRUE}
ggplot(adult, aes (x = SRAGE_P, fill= factor(RBMI))) +
  geom_histogram(aes(y = ..count../sum(..count..)), binwidth = 1, position = "fill") +
  BMI_fill +
  facet_grid(RBMI ~ .)
```

# Create DF with table()

```{r, eval=FALSE, include=TRUE}
DF <- table(adult$RBMI, adult$SRAGE_P)
```

# Use apply on DF to get frequency of each group

```{r, eval=FALSE, include=TRUE}
DF_freq <- apply(DF, 2, function(x) x/sum(x) )
```

# Load reshape2 and use melt on DF to create DF_melted

```{r, eval=FALSE, include=TRUE}
library(reshape2)
DF_melted <- melt(DF_freq)
```

# Change names of DF_melted

```{r, eval=FALSE, include=TRUE}
names(DF_melted) <- c("FILL", "X", "value") 
```


# Add code to make this a faceted plot

```{r, eval=FALSE, include=TRUE}
ggplot(DF_melted, aes(x = X, y = value, fill = FILL)) +
  geom_bar(stat = "identity", position = "stack") +
  BMI_fill +
  facet_grid(FILL ~ .)
```

Part 6: Merimeko/Mosaic Plot

# The initial contingency table

```{r,eval=FALSE, include=TRUE }
DF <- as.data.frame.matrix(table(adult$SRAGE_P, adult$RBMI))
```

# Add the columns groupsSum, xmax and xmin. Remove groupSum again.

```{r, eval=FALSE, include=TRUE}
DF$groupSum <- rowSums(DF)
DF$xmax <- cumsum(DF$groupSum)
DF$xmin <- DF$xmax - DF$groupSum
```

# The groupSum column needs to be removed, don't remove this line

```{r, eval=FALSE, include=TRUE}
DF$groupSum <- NULL
```

# Copy row names to variable X

```{r, eval=FALSE, include=TRUE}
DF$X <- row.names(DF)
```

# Melt the dataset

```{r, eval=FALSE, include=TRUE}
library(reshape2)
DF_melted <- melt(DF, id.vars = c("X", "xmin", "xmax"), variable.name = "FILL")
```

# dplyr call to calculate ymin and ymax - don't change

```{r, eval=FALSE, include=TRUE}
library(dplyr)
DF_melted <- DF_melted %>%
  group_by(X) %>%
  mutate(ymax = cumsum(value/sum(value)),
         ymin = ymax - value/sum(value))
```

# Plot rectangles - don't change.

```{r, eval=FALSE, include=TRUE}
library(ggthemes)
ggplot(DF_melted, aes(ymin = ymin,
                 ymax = ymax,
                 xmin = xmin,
                 xmax = xmax,
                 fill = FILL)) +
  geom_rect(colour = "white") +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  BMI_fill +
  theme_tufte()
```

Part 7: Adding Statistics

# Perform chi.sq test (RBMI and SRAGE_P)

```{r, eval=FALSE, include=TRUE}
results <- chisq.test(table(adult$RBMI, adult$SRAGE_P))
```

# Melt results$residuals and store as resid

```{r, eval=FALSE, include=TRUE}
resid <- melt(results$residuals)
```

# Change names of resid

```{r, eval=FALSE, include=TRUE}
names(resid) <- c("FILL", "X", "residual")
```

# merge the two datasets:

```{r, eval=FALSE, include=TRUE}
DF_all <- merge(DF_melted, resid)
```

# Update plot command

```{r, eval=FALSE, include=TRUE}
library(ggthemes)
ggplot(DF_all, aes(ymin = ymin,
                   ymax = ymax,
                   xmin = xmin,
                   xmax = xmax,
                   fill = residual)) +
  geom_rect() +
  scale_fill_gradient2() +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  theme_tufte()
```

Part 8: Adding Text

# Position for labels on x axis

```{r, eval=FALSE, include=TRUE}
DF_all$xtext <- DF_all$xmin + (DF_all$xmax - DF_all$xmin)/2
```

# Position for labels on y axis (don't change)

```{r, eval=FALSE, include=TRUE}
index <- DF_all$xmax == max(DF_all$xmax)
DF_all$ytext <- DF_all$ymin[index] + (DF_all$ymax[index] - DF_all$ymin[index])/2
```

# Plot

```{r, eval=FALSE, include=TRUE}
ggplot(DF_all, aes(ymin = ymin, ymax = ymax, xmin = xmin,
                   xmax = xmax, fill = residual)) +
  geom_rect(col = "white") +
  # geom_text for ages (i.e. the x axis)
  geom_text(aes(x = xtext,
                label = X),
            y = 1,
            size = 3,
            angle = 90,
            hjust = 1,
            show.legend = FALSE) +
  # geom_text for BMI (i.e. the fill axis)
  geom_text(aes(x = max(xmax),
                y = ytext,
                label = FILL),
            size = 3,
            hjust = 1,
            show.legend  = FALSE) +
  scale_fill_gradient2() +
  theme_tufte() +
  theme(legend.position = "bottom")
```

Part 9: Generalizations

# Load all packages

```{r, eval=FALSE, include=TRUE}
library(ggplot2)
library(reshape2)
library(dplyr)
library(ggthemes)
```

# Script generalized into a function

```{r, eval=FALSE, include=TRUE}
mosaicGG <- function(data, X, FILL)  
```

# Proportions in raw data

```{r, eval=FALSE, include=TRUE}
  DF <- as.data.frame.matrix(table(data[[X]], data[[FILL]]))
  DF$groupSum <- rowSums(DF)
  DF$xmax <- cumsum(DF$groupSum)
  DF$xmin <- DF$xmax - DF$groupSum
  DF$X <- row.names(DF)
  DF$groupSum <- NULL
  DF_melted <- melt(DF, id = c("X", "xmin", "xmax"), variable.name = "FILL")
  DF_melted <- DF_melted %>%
    group_by(X) %>%
    mutate(ymax = cumsum(value/sum(value)),
           ymin = ymax - value/sum(value))

```

# Chi-sq test
 
```{r, eval=FALSE, include=TRUE}
 results <- chisq.test(table(data[[FILL]], data[[X]])) # fill and then x
  resid <- melt(results$residuals)
  names(resid) <- c("FILL", "X", "residual")
```

# Merge data

```{r, eval=FALSE, include=TRUE}
  DF_all <- merge(DF_melted, resid)
```

# Positions for labels

```{r, eval=FALSE, include=TRUE}
  DF_all$xtext <- DF_all$xmin + (DF_all$xmax - DF_all$xmin)/2
  index <- DF_all$xmax == max(DF_all$xmax)
  DF_all$ytext <- DF_all$ymin[index] + (DF_all$ymax[index] - DF_all$ymin[index])/2
```

# plot:

```{r, eval=FALSE, include=TRUE}
  g <- ggplot(DF_all, aes(ymin = ymin,  ymax = ymax, xmin = xmin,
                          xmax = xmax, fill = residual)) +
  geom_rect(col = "white") +
  geom_text(aes(x = xtext, label = X),
            y = 1, size = 3, angle = 90, hjust = 1, show.legend = FALSE) +
  geom_text(aes(x = max(xmax),  y = ytext, label = FILL),
            size = 3, hjust = 1, show.legend = FALSE) +
  scale_fill_gradient2("Residuals") +
  scale_x_continuous("Individuals", expand = c(0,0)) +
  scale_y_continuous("Proportion", expand = c(0,0)) +
  theme_tufte() +
  theme(legend.position = "bottom")
  print(g)
```

# BMI described by age

```{r, eval=FALSE, include=TRUE}
mosaicGG(adult, X = "SRAGE_P", FILL = "RBMI")
```

# Poverty described by age

```{r, eval=FALSE, include=TRUE}
mosaicGG(adult, X = "SRAGE_P", FILL = "POVLL")
```

# mtcars: am described by cyl

```{r, eval=FALSE, include=TRUE}
mosaicGG(mtcars, X = "cyl", FILL = "am")
```

# Vocab: vocabulary described by education

```{r, eval=FALSE, include=TRUE}
library(car)
mosaicGG(Vocab, X = "education", FILL = "vocabulary")
```

